Automatic word stress annotation of Russian unrestricted text

### 2021-03-28 2015-rusaccent-nodalida/rusaccent.tex
Rename; Backup

\usepackage{times}
\usepackage{latexsym}
\usepackage{fixltx2e} %allows subscripts
%\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

\begin{document}
\maketitle
\begin{abstract}
  We evaluate the effectiveness of finite-state tools we developed for 
  automatically annotating word stress in Russian unrestricted text. This task is
  relevant for computer-assisted language learning and text-to-speech. To our 
  knowledge, this is the first study to empirically evaluate the results of this 
  task. Given an adequate lexicon with specified stress, the primary obstacle for 
  correct stress placement is disambiguating homographic wordforms. The baseline
  performance of this task is 90.07\%, (known words only, no morphosyntactic 
  disambiguation). Using a Constraint Grammar to disambiguate homographs, we 
  achieve 93.21\% accuracy with minimal errors. For applications with a higher 
  threshold for errors, we achieved 96.15\% accuracy by incorporating frequency-
  based guessing and a simple algorithm for guessing the stress position on 
  unknown words. These results highlight the need for morphosyntactic 
  disambiguation in the wordstress placement task for Russian, and set a standard 
  for future research on this task.
\end{abstract}
\rr{language learning?}

\rr{REV3: You should use "gold-standard" consistently throughout (sometimes the authors
write "gold" to mean "gold-standard", which sounds strange.)}
\rr{Anya: wordstress or word stress?}
\section{Introduction}
Lexical stress and its attendant vowel reduction are a prominent feature of 
spoken Russian; the incorrect placement of stress can render speech almost 
incomprehensible. This is because Russian word stress is phonemic, i.e. many 
wordforms are distinguished from one another only by stress position. 
This is the cause of considerable difficulty for learners, since 
the inflecting word classes include 
complex patterns of shifting stress, and a lexeme's stress pattern cannot be 
predicted from surface forms. Furthermore, standard written Russian does not 
typically mark word stress.\footnote{Texts intended for native speakers sometimes 
mark stress on words that cannot be disambiguated through context. Theoretically, 
a perfect wordstress placement system could help an author identify
tokens which should be stressed for natives: any token that cannot
be disambiguated by syntactic or semantic means should be marked for stress.} 
Without information about lexical stress position, 
correctly converting written Russian text to speech is impossible. This is true 
both for humans (e.g. foreign language students) and computers (e.g. 
text-to-speech).\rr{add something about how TTS systems handle this right now. Just no one evaluates how well they are doing.}

We identify three different types of word stress ambiguity.
\emph{Intraparadigmatic} stress ambiguity refers
to homographic wordforms belonging to the same lexeme, as shown in 
\ref{ex:intrahom}.

\needspace{4\baselineskip} % keeps example all together (no orphan line) 
\ex. Intraparadigmatic homographs \label{ex:intrahom}
\a. \rus{т\'{е}ла} \emph{t\'{e}la} `body.\textsubscript{SG-GEN}' 
    \label{ex:bodySGGEN}
\b. \rus{тел\'{а}} \emph{tel\'{a}} `body.\textsubscript{PL-NOM}' 
    \label{ex:bodyPLNOM}

@@

\ex. Morphosyntactically incongruent homographs \label{ex:MSincongruent}
\a. \rus{н\'{а}шей} \emph{nášej} `our.\textsubscript{F-SG-GEN/DAT/LOC...}'\\
    \rus{наш\'{е}й} \emph{našéj} `sew on.\textsubscript{IMP-2SG}'
\b. \rus{дор\'{о}га} \emph{doróga} `road.\textsubscript{N-F-SG-NOM}'\\
    \rus{дорог\'{а}} \emph{dorogá} `dear.\textsubscript{ADJ-F-SG-PRED}'

\emph{Morphosyntactically congruent} stress ambiguity occurs between homographs 
that belong to separate lexemes, 
and whose morphosyntactic values are identical, as shown in
\ref{ex:MScongruent}. This kind of stress ambiguity is relatively rare, and resolving this ambiguity would require the use of technologies such as Word Sense Disambiguation.
%\rr{cite Tore's book on stress?}

\ex. Morphosyntactically congruent homographs \label{ex:MScongruent}
\a. \rus{з\'{a}мок} \emph{z\'{a}mok} `castle.\textsubscript{SG-NOM}'\\
	\rus{зам\'{о}к} \emph{zam\'{o}k} `lock.\textsubscript{SG-NOM}'
\b. \rus{з\'{a}мков} \emph{z\'{a}mkov} `castle.\textsubscript{PL-GEN}'\\
	\rus{замк\'{о}в} \emph{zamk\'{o}v} `lock.\textsubscript{PL-GEN}'
\c. ...\\
	...
%\begin{table}
%\begin{center}
%\begin{tabular}{c|ll}
%\bf Number-Case & \bf `castle' & \bf `lock' \\
%\hline
%\texttt{SG-NOM} & \rus{з\'{a}мок} & \rus{зам\'{о}к} \\
%\texttt{SG-GEN} & \rus{з\'{a}мка} & \rus{замк\'{а}} \\
%\texttt{PL-NOM} & \rus{з\'{a}мки} & \rus{замк\'{и}} \\
%\texttt{PL-GEN} & \rus{з\'{a}мков} & \rus{замк\'{о}в} \\
%... & \hskip 1em ... & \hskip 1em ... \\
%\end{tabular}
%\end{center}
%\caption{ \label{table:MScongruent} Morphosyntactically congruent homographs. }
%\end{table}

@@

It should be noted that not all morphosyntactic ambiguity leads to stress 
ambiguity. For example, \emph{novyj} `new' (and every other adjective) has 
identical forms for F-SG-GEN, F-SG-LOC, F-SG-DAT and F-SG-INS: \emph{n\'{o}voj}. 
Likewise, the form \emph{vypej} has multiple possible readings 
(including `drink.IMP', `bittern.PL-GEN'), but they all have the same stress 
position: \emph{v\'{y}pej}. We refer to this as \emph{stress-irrelevant} 
morphosyntactic ambiguity, since all readings have the same stress placement.

In the case of unrestricted text in Russian, most stress placement 
ambiguity is rooted in intraparadigmatic and morphosyntactically incongruent 
ambiguity. Detailed part-of-speech tagging with morphosyntactic analysis
can help determine the stress of these forms, since each alternative stress 
placement is tied to a different tag sequence. In this study we focus on 
the role of detailed part-of-speech tagging in improving
automatic stress placement. We leave morphosyntactically congruent stress 
ambiguity to future work because it is by far the least common type of stress
\rr{Exactly how common is congruent stress ambiguity? REVIEW 2: Since this type of ambiguity cannot be resolved with your approach, it imposes an upper bound. So, it would be very interesting to see how many examples of morphosyntactically congruent stress ambiguity you found in the annotated corpus.}
ambiguity, and disambiguating morphosyntactically congruent stress requires 
fundamentally different technology from the other approaches of this study.

\subsection{Background and task definition} \label{background}
\rr{cite diacritic restoration}
Automatic stress placement in Russian is similar to diacritic restoration, a task
which has received increasing interest over the last 20 years. \rr{cite?}Generally 
speaking, diacritics disambiguate otherwise homographic wordforms, so missing 
diacritics can complicate many NLP tasks, such as text-to-speech. For example, 
speakers of Czech may type emails and other communications
without standard diacritics. In order to generate speech from these texts, they 
must first be normalized by restoring diacritics.
\rr{Sasha has citations from a lab in Belorus. Lobanov.}
A slightly different situation arises with languages whose standard orthography 
is underspecified, like vowel quality in Arabic or Hebrew. For such languages, 
the `restoration' of vowel diacritics results in less ambiguity than in standard 
orthography. For languages with inherently ambiguous orthography,
it may be more precise to refer to this as `diacritic enhancement', since it 
produces text that is less ambiguous than the standard language. In this sense, 
Russian orthography is similar to Arabic and Hebrew, since its vowel qualities 
are underspecified in standard orthography. \rr{spell out why stress placement is important for pronunciation.}

Many studies of Russian text-to-speech and automatic speech recognition make note 
of the difficulties caused by the shortcomings of their stress-marking resources 
(e.g. \newcite{krivnova_1998}), but very few studies have targeted stress-marking 
itself, and those that do focus on placing stress on unknown words, with almost 
no attention to the problems that arise with running text. For example, 
\newcite{Xomicevic_2008} developed a set of heuristics for guessing stress 
placement on unknown words in Russian. More recently, 
\newcite{hall_sproat_russian_2013} trained a maximum entropy model on a 
dictionary of Russian words, and evaluated on wordlists containing `known' and 
`unknown' wordforms.\footnote{\newcite{hall_sproat_russian_2013} randomly 
selected their training and test data from a list of wordforms, and so a number 
of lexemes had wordforms in both the training and test data. Wordforms in the 
test data whose sibling wordforms from the same lexeme were in the training set 
were categorised as `known' wordforms.} Their model achieved 98.7\% accuracy on 
known words, and 83.9\% accuracy on unknown words.

\rr{somehow make this point clearer. Machine learning is not possible for stress in running text because large corpora do not exist.}
One major shortcoming of evaluating on wordlists is that they do not reflect 
wordform frequency.
Many of the most problematic stress ambiguities in Russian occur in 
high-frequency wordforms, so evaluations based on wordlists - as opposed to 
running text - have inflated results (see discussion in 
Section~\ref{discussion}). Furthermore, working with running text allows for the
possibility of disambiguating homographs based on syntactic context.
\rr{REVIEWER 3: Why are results "inflated"?
if the system does better on more rare forms, then results are "deflated".  if
it does better on more frequent forms, they are "inflated".   This seems to be
just a different kind of evaluation.
It seems that the point of evaluating on running text ought to be precisely
that the same text could have an ambiguous word with more than one stress
pattern. So this is really evaluating a different task.}

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{r|c|c}
    & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} \\
    & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} \\
    & \rus{костить}-V-IPFV-IMP\hskip 1em\rus{кост\'{и}} & \\
    \hline
    {\small {\tt bare}} & \rus{кости} & \rus{кости} \\
    {\small {\tt safe}} & \rus{кости} & \rus{к\'{о}сти} \\
    {\small {\tt guess}} & \rus{к\'{о}сти} ($p$=0.67) or \rus{кост\'{и}} ($p$=0.33) & \rus{к\'{о}сти}\\
    {\small {\tt guessFreq}} & \rus{к\'{о}сти} & \rus{к\'{о}сти}
  \end{tabular}
  \caption{Example output of each stress placement approach, given a particular set of readings for the token \rus{кости} \emph{kosti}.}
  \label{tab:conditions}
\end{table*}
\subsection{Stress corpus}
\rr{imperative kosti is weird}
\rr{Table 2 needs more explanation: what exactly these two columns mean and how you ended up with these two sets of readings}

Russian texts with marked word stress are relatively rare, except in materials
for language learners, which are predominantly proprietary. Our gold corpus was 
collected from
free texts on Russian language-learning websites.\footnote{URLs can be found in
the X data repository (URL will be added in the final version).} This small
corpus (7689 tokens) is representative of texts 
that learners of Russian are likely to encounter in their studies. These texts 
include excerpts from well-known literary works, as well as dialogs and 
prose and individual sentences that were written for learners. \rr{1st and 2nd language learners? Sometimes children's textbooks have them, especially when syllables are separated.}
\rr{Why not RNC corpus for gold corpus?}

Unfortunately, the general practice for marking stress in Russian is to
\emph{not} mark stress on monosyllabic tokens, effectively assuming that
all monosyllabics are stressed.\rr{citation?} However, this approach 
is not well-motivated. Many words -- both monosyllabic and multisyllabic -- are 
unstressed, especially among prepositions, conjunctions, and particles.
%multisyllabics that can be unstressed: \rus{через, между} 
Furthermore, there are many
\rr{examples}
high-frequency monosyllabic homographs that can be either stressed or unstressed, 
depending on their part of speech, or particular collocations.\rr{examples} Clearly, for
such words, one cannot assume that they are stressed on the basis of their 
syllable count.
\rr{monos are determining WHETHER stress is present. slightly different.}

Based on these considerations, we built our tools to mark stress on every
word, both monosyllabic and multisyllabic. However, because our gold corpus texts 
do not mark stress on monosyllabic words, we cannot evaluate our annotation of 
those words.

Similarly, some compound Russian words have secondary stress, but this is rarely 
marked, if at all, even in educational materials. Therefore, even though our 
tools are built to mark secondary stress, we cannot evaluate secondary stress
marks, since they are absent in our gold corpus.

In order to test our wordstress placement system, we removed all stress marks
from the gold corpus, then marked stress on the unstressed version using our tools, and then
compared with the original.

@@

State-of-the-art morphological analysis in Russian is based on finite-state
technology \cite{Nozhov-03,Segalovich-03}. To our knowledge, no existing 
open-source resources are available for analyzing and generating 
stressed wordforms. Therefore, we developed free and open-source finite-state 
tools capable of analyzing and generating stressed wordforms, based on the
well-known \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
Our Finite-State Transducer\footnote{Using two-level 
morphology \cite{Koskenniemi-84}, implemented in both xfst
\cite{Beesley.Karttunen-03} and hfst \cite{hfst-11}} (FST) generates all possible 
morphosyntactic readings of each wordform, and our Constraint 
Grammar\footnote{Implemented using vislcg3 constraint grammar parser
(http://beta.visl.sdu.dk/cg3.html).}
\cite{Karlsson-90,Karlsson.Voutilainen.ea-95} then removes
some readings based on syntactic context. The ultimate success of our 
stress placement system depends on the performance of the Constraint Grammar. 
Ideally, the Constraint Grammar would successfully remove all but the correct 
reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described below.

\rr{rename guess > guessRand}
\rr{more about our corpus}
\rr{explain guess better}
The {\small {\tt bare}} approach is to not mark stress on words with more than 
one reading. Since both sets of readings in Table~\ref{tab:conditions} have more
than one reading, {\small {\tt bare}} does not output a stressed form.

@@

The {\small {\tt guess}} approach is to randomly select one of the available 
readings. In the first column of
Table~\ref{tab:conditions}, a random selection means that \rus{к\'{о}сти} 
\emph{k\'{o}sti} is twice as likely as \rus{кост\'{и}} \emph{kost\'{i}}. The
second column of Table~\ref{tab:conditions} contains stress-irrelevant
ambiguity, so a random selection of a reading has the same result as the
{\small {\tt safe}} approach.

The {\small {\tt guessFreq}} 
approach is to select the reading that is most frequent, with frequency data
taken from a separate hand-disambiguated corpus. If none of the readings are 
found in the corpus, then {\small {\tt guessFreq}} selects the reading
with the tag sequence (lemma removed) that is most frequent in our corpus. If the 
tag sequence is not found
in our frequency list, then {\small {\tt guessFreq}} backs off to the 
{\small {\tt guess}} algorithm. In the first column of
Table~\ref{tab:conditions}, {\small {\tt guessFreq}} selects \rus{к\'{о}сти} 
\emph{k\'{o}sti} because the tag sequence N-F-SG-GEN is more frequent than the
other alternatives. Note 
that for tokens with stress-irrelevant ambiguity (e.g. the second column of
Table~\ref{tab:conditions}), {\small {\tt guess}} and 
{\small {\tt guessFreq}} produce the same result as the {\small {\tt safe}} 
method.

So far, the approaches discussed are dependent on the availability of readings 
from the FST. The focus of our study is on disambiguation of known 
words, but we also wanted to guess the stress of unknown tokens in order to 
establish some kind of accuracy maximum for applications that are more tolerant 
of higher error rates. To this end, we selected a simple guessing method for 
unknown words. A recent study by \newcite{Lavitskaya.Kabak-14} concludes that 
Russian has default final stress in consonant-final words, and penultimate stress 
in vowel-final words.\footnote{There is some controversy over how to define 
default stress in Russian. \newcite{Crosswhite.ea-03} conclude that the default 
stress position in Russian is the last syllable of the stem.} Based on this 
\rr{what is the difference between last syllable and last vowel followed by consonant?}
conclusion, our syllable-guessing method places the stress on the last vowel that 
is followed by a consonant.\footnote{Although this approach is simplistic, 
unknown words are not the central focus 
of this study. More sophisticated heuristics and machine-
learning approaches to unknown words are discussed in Section~\ref{discussion}.} This method 
is applied in two approaches, {\small {\tt guessSyll}} 
and {\small {\tt guessFreqSyll}}, which are otherwise identical to 
{\small {\tt guess}} and {\small {\tt guessFreq}}, respectively.
\rr{REVIEWER 2: However, one can imagine a token that gets some morphological readings, none of them being correct (for example, when the correct lexeme is not covered by the dictionary -- but an homonymous incorrect lexeme is). Do you have any estimations of how often this happens in the running text?}

For our baseline, we take the output of our morphological analyzer (without the 
Constraint Grammar) in combination with the {\small {\tt bare}}, 
{\small {\tt safe}}, {\small {\tt guess}}, {\small {\tt guessFreq}}, 
{\small {\tt guessSyll}}, and {\small {\tt guessFreqSyll}} approaches. We also 
compare our outcomes with the RussianGram\footnote{http://russiangram.com/} 
plugin for the Google Chrome web browser. RussianGram is not open-source, so we
can only guess what technologies support the service. In any case, it provides a
meaningful reference point for the success of each of the methods described 
above.

>>>>

\usepackage{times}
\usepackage{latexsym}
\usepackage{fixltx2e} %allows subscripts
\usepackage{tipa} %allows IPA symbols
%\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

\begin{document}
\maketitle
\begin{abstract}
  We evaluate the effectiveness of finite-state tools we developed for 
  automatically annotating word stress in Russian unrestricted text. This task is
  relevant for computer-assisted language learning and text-to-speech. To our 
  knowledge, this is the first study to empirically evaluate the results of this 
  task. Given an adequate lexicon with specified stress, the primary obstacle for 
  correct stress placement is disambiguating homographic wordforms. The baseline
  performance of this task is 90.07\%, (known words only, no morphosyntactic 
  disambiguation). Using a Constraint Grammar to disambiguate homographs, we 
  achieve 93.21\% accuracy with minimal errors. For applications with a higher 
  threshold for errors, we achieved 96.15\% accuracy by incorporating frequency-
  based guessing and a simple algorithm for guessing the stress position on 
  unknown words. These results highlight the need for morphosyntactic 
  disambiguation in the word stress placement task for Russian, and set a standard 
  for future research on this task.
\end{abstract}
\rr{language learning?}

\section{Introduction}

Lexical stress and its attendant vowel reduction are a prominent feature of 
spoken Russian; the incorrect placement of stress can render speech almost 
incomprehensible. This is because Russian word stress is phonemic, i.e. many 
wordforms are distinguished from one another only by stress position. 
This is the cause of considerable difficulty for learners, since 
the inflecting word classes include 
complex patterns of shifting stress, and a lexeme's stress pattern cannot be 
predicted from surface forms. Furthermore, standard written Russian does not 
typically mark word stress.\footnote{Texts intended for native speakers sometimes 
mark stress on words that cannot be disambiguated through context. Theoretically, 
a perfect word stress placement system could help an author identify
tokens which should be stressed for natives: any token that cannot
be disambiguated by syntactic or semantic means should be marked for stress.} 
Without information about lexical stress position, 
correctly converting written Russian text to speech is impossible. Half of the
vowel letters in Russian change their pronunciation significantly, depending on
their position relative to the stress. For example the word \emph{dogovórom}
`contract.SG-INS' is pronounced as /\textipa{d@g2vOr@m}/. Determining such vowel
qualities is impossible without specifying the stress position. This is a problem 
both for humans (e.g. foreign language students) and computers (e.g. 
text-to-speech).

We identify three different types relations between word stress ambiguity and
morphosyntactic ambiguity. First, \emph{intraparadigmatic} stress ambiguity 
refers to homographic wordforms belonging to the same lexeme, as shown in 
\ref{ex:intrahom}.\footnote{Cyrillic is transliterated using the scientific transliteration
scheme.}

\needspace{4\baselineskip} % keeps example all together (no orphan line) 
\ex. Intraparadigmatic homographs \label{ex:intrahom}
\a. \emph{t\'{e}la} `body.\textsubscript{SG-GEN}' 
    \label{ex:bodySGGEN}
\b. \emph{tel\'{a}} `body.\textsubscript{PL-NOM}' 
    \label{ex:bodyPLNOM}

@@

\ex. Morphosyntactically incongruent homographs \label{ex:MSincongruent}
\a. \emph{nášej} `our.\textsubscript{F-SG-GEN/LOC/DAT/INS}'\\
    \emph{našéj} `sew on.\textsubscript{IMP-2SG}'
\b. \emph{doróga} `road.\textsubscript{N-F-SG-NOM}'\\
    \emph{dorogá} `dear.\textsubscript{ADJ-F-SG-PRED}'

\emph{Morphosyntactically congruent} stress ambiguity occurs between homographs 
that belong to separate lexemes, 
and whose morphosyntactic values are identical, as shown in
\ref{ex:MScongruent}. This kind of stress ambiguity is relatively rare, 
and resolving this ambiguity would require the use of technologies such as Word Sense Disambiguation.

\ex. Morphosyntactically congruent homographs \label{ex:MScongruent}
\a. \emph{z\'{a}mok} `castle.\textsubscript{SG-NOM}'\\
	\emph{zam\'{o}k} `lock.\textsubscript{SG-NOM}'
\b. \emph{z\'{a}mkov} `castle.\textsubscript{PL-GEN}'\\
	\emph{zamk\'{o}v} `lock.\textsubscript{PL-GEN}'
\c. ...\\
	...
%\begin{table}
%\begin{center}
%\begin{tabular}{c|ll}
%\bf Number-Case & \bf `castle' & \bf `lock' \\
%\hline
%\texttt{SG-NOM} & \rus{з\'{a}мок} & \rus{зам\'{о}к} \\
%\texttt{SG-GEN} & \rus{з\'{a}мка} & \rus{замк\'{а}} \\
%\texttt{PL-NOM} & \rus{з\'{a}мки} & \rus{замк\'{и}} \\
%\texttt{PL-GEN} & \rus{з\'{a}мков} & \rus{замк\'{о}в} \\
%... & \hskip 1em ... & \hskip 1em ... \\
%\end{tabular}
%\end{center}
%\caption{ \label{table:MScongruent} Morphosyntactically congruent homographs. }
%\end{table}

@@

It should be noted that most morphosyntactic ambiguity in unrestricted text
does not result in stress ambiguity. For example, \emph{novyj} `new' (and every 
other adjective) has identical forms for F-SG-GEN, F-SG-LOC, F-SG-DAT and F-SG-
INS: \emph{n\'{o}voj}. Likewise, the form \emph{vypej} has multiple possible 
readings (including `drink.IMP', `bittern.PL-GEN'), but they all have the same 
stress position: \emph{v\'{y}pej}. We refer to this as \emph{stress-irrelevant} 
morphosyntactic ambiguity, since all readings have the same stress placement.

In the case of unrestricted text in Russian, most stress placement 
ambiguity is rooted in intraparadigmatic and morphosyntactically incongruent 
ambiguity. Detailed part-of-speech tagging with morphosyntactic analysis
can help determine the stress of these forms, since each alternative stress 
placement is tied to a different tag sequence. In this study we focus on 
the role of detailed part-of-speech tagging in improving
automatic stress placement. We leave morphosyntactically congruent stress 
ambiguity to future work because it is by far the least common type of stress
ambiguity ($\approx$~1.2\% of tokens in running text), and disambiguating morphosyntactically congruent stress requires 
fundamentally different technology from the other approaches of this study.
%\rr{Exactly how common is congruent stress ambiguity? REVIEW 2: Since this type of ambiguity cannot be resolved with your approach, it imposes an upper bound. So, it would be very interesting to see how many examples of morphosyntactically congruent stress ambiguity you found in the annotated corpus.}

\subsection{Background and task definition} \label{background}
%\rr{cite diacritic restoration}
Automatic stress placement in Russian is similar to diacritic restoration, a task
which has received increasing interest over the last 20 years. \rr{cite?}Generally 
speaking, diacritics disambiguate otherwise homographic wordforms, so missing 
diacritics can complicate many NLP tasks, such as text-to-speech. For example, 
speakers of Czech may type emails and other communications
without standard diacritics. In order to generate speech from these texts, they 
must first be normalized by restoring diacritics.
%\rr{Sasha has citations from a lab in Belorus. Lobanov.}
A slightly different situation arises with languages whose standard orthography 
is underspecified, like vowel quality in Arabic or Hebrew. For such languages, 
the `restoration' of vowel diacritics results in less ambiguity than in standard 
orthography. For languages with inherently ambiguous orthography,
it may be more precise to refer to this as `diacritic enhancement', since it 
produces text that is less ambiguous than the standard language. In this sense, 
Russian orthography is similar to Arabic and Hebrew, since its vowel qualities 
are underspecified in standard orthography.

\rr{add something about how TTS systems handle this right now. Just no one evaluates how well they are doing.}
Many studies of Russian text-to-speech and automatic speech recognition make note 
of the difficulties caused by the shortcomings of their stress-marking resources 
(e.g. \newcite{krivnova-98}), but very few studies have targeted stress-marking 
itself, and those that do focus on placing stress on unknown words, with almost 
no attention to the problems that arise with known words in running text. For example, 
\newcite{Xomicevic.Rybin.ea-08} developed a set of heuristics for guessing stress 
placement on unknown words in Russian. More recently, 
\newcite{Hall.Sproat-13} trained a maximum entropy model on a 
dictionary of Russian words, and evaluated on wordlists containing `known' and 
`unknown' wordforms.\footnote{\newcite{Hall.Sproat-13} randomly 
selected their training and test data from a list of wordforms, and so a number 
of lexemes had wordforms in both the training and test data. Wordforms in the 
test data whose sibling wordforms from the same lexeme were in the training set 
were categorized as `known' wordforms.} Their model achieved 98.7\% accuracy on 
known words, and 83.9\% accuracy on unknown words. The task of training and 
evaluating on wordlists is different from that of placing
stress in running text, so these results are not comparable to the results reported
in this paper. Since many of the most problematic stress ambiguities in 
Russian occur in high-frequency wordforms, evaluations of wordform lists encounter
stress ambiguity seven times less frequently than in running text (see discussion in 
Section~\ref{discussion}). Furthermore, working with running text includes the
possibility of disambiguating homographs based on syntactic context.

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{r|c|c}
%    & Reading set 1 & Reading set 2 \\
%    \hline
   Readings: & \rus{кость}-N-F-SG-GEN\hskip 1em\emph{kósti} & \rus{кость}-N-F-SG-GEN\hskip 1em\emph{kósti} \\
    & \rus{кость}-N-F-SG-DAT\hskip 1em\emph{kósti} & \rus{кость}-N-F-SG-DAT\hskip 1em\emph{kósti} \\
    & \rus{костить}-V-IPFV-IMP~~\emph{kostí} & \\
    \hline
    {\small {\tt bare}} & \emph{kosti} & \emph{kosti} \\
    {\small {\tt safe}} & \emph{kosti} & \emph{kósti} \\
    {\small {\tt randReading}} & \emph{kósti} ($p$=0.67) or \emph{kostí} ($p$=0.33) & \emph{kósti}\\
    {\small {\tt freqReading}} & \emph{kósti} & \emph{kósti}
  \end{tabular}
  \caption{Example output of each stress placement approach, given a particular set of readings for the token \emph{kosti}}
  \label{tab:conditions}
\end{table*}

\subsection{Stress corpus}

Russian texts with marked word stress are relatively rare, except in materials
for second language learners, which are predominantly proprietary. Our gold-standard corpus was 
collected from
free texts on Russian language-learning websites.\footnote{URLs can be found in
the X data repository (URL will be added in the final version).} This small
\rr{ADD TO TROLLING!}
corpus (7689 tokens) is representative of texts 
that learners of Russian are likely to encounter in their studies. These texts 
include excerpts from well-known literary works, as well as dialogs, 
prose, and individual sentences that were written for learners. 
%\rr{Why not RNC corpus for gold-standard corpus?}

Unfortunately, the general practice for marking stress in Russian is to
\emph{not} mark stress on monosyllabic tokens, effectively assuming that
all monosyllabics are stressed. However, this approach 
is not well-motivated. Many words -- both monosyllabic and multisyllabic -- are 
unstressed, especially among prepositions, conjunctions, and particles.
%multisyllabics that can be unstressed: \rus{через, между} 
Furthermore, there are many
high-frequency monosyllabic homographs that can be either stressed or unstressed, 
depending on their part of speech, or particular collocations. For example,
the token \emph{čto} is stressed when it means `what' and unstressed in the
conjunction \emph{potomu čto} `because'. For such words, one cannot simply
assume that they are stressed on the basis of their syllable count.
%\rr{monos are determining WHETHER stress is present. slightly different. This is very relevant for syllable-guessing algorithms.}

Based on these considerations, we built our tools to mark stress on every
word, both monosyllabic and multisyllabic. However, because our gold-standard corpus texts 
do not mark stress on monosyllabic words, we cannot evaluate our annotation of 
those words.

Similarly, some compound Russian words have secondary stress, but this is rarely 
marked, if at all, even in educational materials. Therefore, even though our 
tools are built to mark secondary stress, we cannot evaluate secondary stress
marks, since they are absent in our gold-standard corpus.

In order to test our word stress placement system, we removed all stress marks
from the gold-standard corpus, then marked stress on the unstressed version using our tools, and then
compared with the original.

@@

State-of-the-art morphological analysis in Russian is based on finite-state
technology \cite{Nozhov-03,Segalovich-03}. To our knowledge, no existing 
open-source resources are available for analyzing and generating 
stressed wordforms. Therefore, we developed free and open-source finite-state 
tools capable of analyzing and generating stressed wordforms, based on the
well-known \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
Our Finite-State Transducer\footnote{Using two-level 
morphology \cite{Koskenniemi-83,Koskenniemi-84}, implemented in both xfst
\cite{Beesley.Karttunen-03} and hfst \cite{Linden.Silfverberg.ea-11}} (FST) generates all possible 
morphosyntactic readings of each wordform, and our Constraint 
Grammar\footnote{Implemented using vislcg3 constraint grammar parser
(http://beta.visl.sdu.dk/cg3.html).}
\cite{Karlsson-90,Karlsson.Voutilainen.ea-95} then removes
some readings based on syntactic context. The ultimate success of our 
stress placement system depends on the performance of the Constraint Grammar. 
Ideally, the Constraint Grammar would successfully remove all but the correct 
reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described below. Table~\ref{tab:conditions} shows two possible sets of readings
for the token \emph{kosti}, as well as the output of each approach described below.
The first column exhibits stress ambiguity between the noun readings and the imperative
verb reading. The second column shows a similar set of readings, after the
constraint grammar has removed the imperative verb reading. This results in only
stress-irrelevant ambiguity.

\rr{more about our corpus}
The {\small {\tt bare}} approach is to not mark stress on words with more than 
one reading. Since both sets of readings in Table~\ref{tab:conditions} have more
than one reading, {\small {\tt bare}} does not output a stressed form.

@@

The {\small {\tt randReading}} approach is to randomly select one of the available 
readings. In the first column of
Table~\ref{tab:conditions}, a random selection means that \emph{kósti} 
is twice as likely as \emph{kostí}. The
second column of Table~\ref{tab:conditions} contains stress-irrelevant
ambiguity, so a random selection of a reading has the same result as the
{\small {\tt safe}} approach.

The {\small {\tt freqReading}} 
approach is to select the reading that is most frequent, with frequency data
taken from a separate hand-disambiguated corpus. If none of the readings are 
found in the corpus, then {\small {\tt freqReading}} selects the reading
with the tag sequence (lemma removed) that is most frequent in our corpus. If the 
tag sequence is not found
in our frequency list, then {\small {\tt freqReading}} backs off to the 
{\small {\tt randReading}} algorithm. In the first column of
Table~\ref{tab:conditions}, {\small {\tt freqReading}} selects \emph{kósti} 
because the tag sequence N-F-SG-GEN is more frequent than the
other alternatives. Note 
that for tokens with stress-irrelevant ambiguity (e.g. the second column of
Table~\ref{tab:conditions}), {\small {\tt randReading}} and 
{\small {\tt freqReading}} produce the same result as the {\small {\tt safe}} 
method.

So far, the approaches discussed are dependent on the availability of readings 
from the FST. The focus of our study is on disambiguation of known 
words, but we also wanted to guess the stress of unknown tokens in order to 
establish some kind of accuracy maximum for applications that are more tolerant 
of higher error rates. To this end, we selected a simple guessing method for 
unknown words. A recent study by \newcite{Lavitskaya.Kabak-14} concludes that 
Russian has default final stress in consonant-final words, and penultimate stress 
in vowel-final words.\footnote{There is some controversy over how to define 
default stress in Russian, cf. \newcite{Crosswhite.Alderete.ea-03}.} Based on this 
%\rr{what is the difference between last syllable and last vowel followed by consonant?}
conclusion, the {\tt guessSyll} method places the stress on the last vowel that 
is followed by a consonant.\footnote{Although this approach is simplistic, 
unknown words are not the central focus 
of this study. More sophisticated heuristics and machine-
learning approaches to unknown words are discussed in Section~\ref{discussion}.} This method 
is applied in two approaches, {\small {\tt randReading+guessSyll}} 
and {\small {\tt freqReading+guessSyll}}, which are otherwise identical to 
{\small {\tt randReading}} and {\small {\tt freqReading}}, respectively.

For our baseline, we take the output of our morphological analyzer (without the 
Constraint Grammar) in combination with the {\small {\tt bare}}, 
{\small {\tt safe}}, {\small {\tt randReading}}, {\small {\tt freqReading}}, 
{\small {\tt randReading+guessSyll}}, and {\small {\tt freqReading+guessSyll}} approaches. We also 
compare our outcomes with the RussianGram\footnote{http://russiangram.com/} 
plugin for the Google Chrome web browser. RussianGram is not open-source, so we
can only guess what technologies support the service. In any case, it provides a
meaningful reference point for the success of each of the methods described 
above.
